Liveness-Based Garbage Collection for Lazy Languages

### 2015-05-18 paper/paper.tex

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves). 
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.45\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\RETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\RETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\RETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%} 
    \end{boxedminipage}

@@

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_6$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live,  since, during execution, \py\ is yet to
  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
  consulted by the garbage collector at this point.
\item In an eager language, \py\ would not have been live at $\pi_8$,
  a program point  beyond its last use.  In  a lazy language, however,
  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
  evaluation  actually takes  place while  evaluating $(\RETURN~\pz)$.
  We therefore  record  the liveness of $\py$ at the program  
  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
  the beginning  of the program $\pi_1$ (since  we do not  kill liveness).
\end{enumerate}

We thus regard  the body of the function as a  tree and consider paths
from  the  root   of  this  tree  to  program   points  which  trigger
evaluation\footnote{Program   points  which  trigger   evaluation  are
  labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
    day}}---\SIF\ conditions  and \RETURN\ expressions.   We call such
paths {\em e-paths}.   For the function \length, there  are three such
paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.

@@

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $(\RETURN~\psi:x)$  with a  demand $\sigma$  forces an
evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
added  to the  union.  Also  notice a  consequences  of laziness---the
entire  expression including  the condition  is not  evaluated  if the
demand on it is $\emptyset$.  This results in the empty map, which, by
abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{ep} \mapsto  \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{ep}$ is a new e-path
  terminating with $\psi$}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in \mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\  
 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
  of  an   expression  $e$.   The  cases  $\RETURN$   and  $\SIF$  are
  straightforward, but note the liveness $x.\epsilon$ generated by the
  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
  three-address instruction:  the liveness of  $e$ is given  by taking
  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
  any incremental  liveness from  $s$.  The main  subtlety is  how the
  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.

%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}

% \end{itemize}

@@

%------------------------------------------------------------%

>>>>

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.

Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves). 
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.45\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%} 
    \end{boxedminipage}

@@

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_6$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live,  since, during execution, \py\ is yet to
  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
  consulted by the garbage collector at this point.
\item In an eager language, \py\ would not have been live at $\pi_8$,
  a program point  beyond its last use.  In  a lazy language, however,
  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
  evaluation  actually takes  place while  evaluating $(\SRETURN~\pz)$.
  We therefore  record  the liveness of $\py$ at the program  
  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
  the beginning  of the program $\pi_1$ (since  we do not  kill liveness).
\end{enumerate}

We thus regard  the body of the function as a  tree and consider paths
from  the  root   of  this  tree  to  program   points  which  trigger
evaluation\footnote{Program   points  which  trigger   evaluation  are
  labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
    day}}---\SIF\ conditions  and \SRETURN\ expressions.   We call such
paths {\em e-paths}.   For the function \length, there  are three such
paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.

@@

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
The  expression $(\SRETURN~\psi:x)$  with a  demand $\sigma$  forces an
evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
added  to the  union.  Also  notice a  consequences  of laziness---the
entire  expression including  the condition  is not  evaluated  if the
demand on it is $\emptyset$.  This results in the empty map, which, by
abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.

%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{ep} \mapsto  \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{ep}$ is a new e-path
  terminating with $\psi$}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in \mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\  
 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
  of  an   expression  $e$.   The  cases  $\SRETURN$   and  $\SIF$  are
  straightforward, but note the liveness $x.\epsilon$ generated by the
  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
  three-address instruction:  the liveness of  $e$ is given  by taking
  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
  any incremental  liveness from  $s$.  The main  subtlety is  how the
  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.

@@

%------------------------------------------------------------%
%\input{lazy-liveness}
